%%-*-latex-*-

\documentclass[11pt,a4paper]{article}

\input{trace}

\title{Quiz \#2 of introduction to networking} 
\author{Christian Rinderknecht} 
\date{\today}

\begin{document}

\maketitle

\begin{center}
\textbf{Duration: one hour and half. Documents are forbidden.}

\textbf{Answers in English will receive extra points.}

\textbf{No Konglish!}
\end{center}


\section{Review questions}

\begin{enumerate}

  \item Why is it that \textsc{ftp} sends control information
    out-of-band?

  \item Why \textsc{ftp} is said to be a stateful protocol?

  \item Are the objectives of flow control and congestion control the
    same?

  \item Give a very short description of how the connection-oriented
    service of the Internet provides reliable transport.

  \item Describe briefly Web caching at proxies and browsers. Why are
    they useful?

  \item Compare the common and different features of \textsc{smtp} and
    \textsc{http}.

  \item What are the components needed to make (use of) cookies?

  \item What is the conditional \texttt{GET} \textsc{http} request
    useful for?

  \item List five tasks that a protocol layer can perform. Is it
    possible that one (or more) of these tasks could be performed by
    two (or more) layers?

  \item What information is used by a process running on one host to
    identify a process running on another host?

\end{enumerate}

\section{Problems}

\begin{enumerate}

  \item Consider sending a file of \(M \times L\) bits over a path of
    \(Q\) links. Each link transmits at \(R\) bits per second. The
    network is lightly loaded so that there are no queuing
    delays. When a form of packet switching is used, the \(M \times
    L\) bits are broken up into \(M\) packets, each packet with \(L\)
    bits. Propagation delay is negligible.
    \begin{enumerate}

      \item Suppose the network is a packet-switched virtual circuit
        network. Denote the VC set-up time by \(t_s\) seconds. Suppose
        the sending layers add a total of \(h\) bits of header to each
        packet. How long does it take to send the file from source to
        destination?

      \item \label{datagram} Suppose the network is a packet-switched
        datagram network and a connectionless service is used. Now
        suppose each packet has \(2h\) bits of header. How long does
        it take to send the file?

      \item Repeat case~\ref{datagram} but assume message switching is
        used (that is, \(2h\) bits are added to the message, and the
        message is not segmented).

      \item Finally, suppose that the network is a circuit-switched
        network. Further suppose that the transmission rate of the
        circuit between source and destination is \(R\)
        bit/s. Assuming \(t_s\) seconds of set-up and \(h\) bits of
        header appended to the entire file, how long does it take to
        send the file?

    \end{enumerate}


\end{enumerate}

\end{document}
