%%-*-latex-*-

\documentclass[12pt,a4paper]{article}

\input{trace}

\title{Mid-term examination on Introduction to the Internet}
\author{Christian Rinderknecht}
\date{\today}

\begin{document}

\maketitle

\section{Review questions}

\begin{enumerate}

  \item What advantage does a circuit-switched network have over a
    packet-switched network? What advantage does TDM have over FDM in
    a circuit-switched network?

  \item Why is it that packet switching is said to employ statistical
    multiplexing? Contrast statistical multiplexing with the
    multiplexing that takes place in TDM.

  \item What is meant by connection state information in a virtual
    circuit network? 

  \item Suppose you are developing a standard for a new type of
    network. You need to decide whether your network will use VCs or
    datagram routing. What are the pros and cons for using VCs?

  \item What are the advantages of message segmentation in
    packet-switched networks? What are the disadvantages?

  \item What is the key distinguishing difference between a tier-1 ISP
    (backbone) and a tier-2 ISP?

  \item Is HFC bandwidth dedicated or shared among users? Are
    collisions possible in a downstream HFC channel?

  \item Consider sending a series of packets from a sending host to a
    receiving host over a fixed route. List the delay components in
    the end-to-end delay for a single packet. Which of these delays
    are constant and which are variable?

  \item List five tasks that a protocol layer can perform. Is it
    possible that one (or more) of these tasks could be performed by
    two (or more) layers?
 
  \item What are the five layers in the Internet protocol stack? What
    are the principal responsibilities of each of these layers?

  \item What information is used by a process running on one host to
    identify a process running on another host?

  \item What is the difference between persistent \textsc{http}
    with pipelining and persistent \textsc{http} without pipelining?

\end{enumerate}

\section{Problems}

\begin{enumerate}

  \item True or false?
  \begin{enumerate}

    \item Suppose a user requests a Web page that consists of some text
      and two images. For this page the client will send one request
      and receive three response messages.

    \item Two distinct Web pages can be sent over the same persistent
      connection.

    \item With non-persistent connections between browser and origin
      ser\-ver, it is possible for a single TCP segment to carry two
      distinct \textsc{http} request messages.

    \item The \verb+Date:+ header in the \textsc{http} response
      message indicates when the object in the response was last
      modified.
   
  \end{enumerate}

  \item Consider sending a file of \(M \times L\) bits over a path of
    \(Q\) links. Each link transmits at \(R\) bits per second. The
    network is lightly loaded so that there are no queuing
    delays. When a form of packet switching is used, the \(M \times
    L\) bits are broken up into \(M\) packets, each packet with \(L\)
    bits. Propagation delay is negligible.
    \begin{enumerate}

      \item Suppose the network is a packet-switched virtual circuit
        network. Denote the VC set-up time by \(t_s\) seconds. Suppose
        the sending layers add a total of \(h\) bits of header to each
        packet. How long does it take to send the file from source to
        destination?

      \item \label{datagram} Suppose the network is a packet-switched
        datagram network and a connectionless service is used. Now
        suppose each packet has \(2h\) bits of header. How long does
        it take to send the file?

      \item Repeat case~\ref{datagram} but assume message switching is
        used (that is, \(2h\) bits are added to the message, and the
        message is not segmented).

      \item Finally, suppose that the network is a circuit-switched
        network. Further suppose that the transmission rate of the
        circuit between source and destination is \(R\)
        bit/s. Assuming \(t_s\) seconds of set-up and \(h\) bits of
        header appended to the entire file, how long does it take to
        send the file?

    \end{enumerate}

  \item Consider sending a large file of \(F\) bits from host A to
    host B. There are two links (and one switch) between them and the
    links are uncongested (that is, no queuing delays). Host A
    segments the file into segments of \(S\) bits each and adds 40
    bits of header to each segment, forming packets of \(L = 40 + S\)
    bits. Each link has a transmission rate of \(R\) bit/s. Assuming
    that \(F/S\) is an integer, find the value of \(S\) that minimises
    the delay of moving the file from host A to host B. Disregard
    propagation delay.

\end{enumerate}

\end{document}
